Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-03-26 paper/intro.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@479 3fc257bb-8550-0410-81b3-a5c103da83f7

\section{Introduction}

  what is MR
  why implement in MPI
  our motivation = graph algorithms, large data sets
    but lib can also be used for any MR algorithm, data agnostic
  novel features of our MR implementation
    C++ lib, open-source, MPI, out-of-core
    no fault tolerance

>>>>

\section{Introduction}
\label{sec:intro}

MapReduce is the programming paradigm popularized by Google
researchers Dean and Ghemawat \cite{Dean}.  Their motivation was to
enable rapid development and deployment of analysis programs to
operate on massive data sets residing on Google's large distributed
clusters.  They introduced a novel way of thinking about certain kinds
of large-scale computations as a "map" operation followed by a
"reduce" operation.  The power of the paradigm is that when cast in
this way, a nominally serial algorithm now becomes two highly parallel
operations working on data local to each processor, sandwiched around
an intermediate data-shuffling operation that requires inter-processor
communication.  The user need only write serial code for the
application-specific map and reduce functions; the parallel data
shuffle can be encapsulated in a library since its operation is
independent of the application.

The Google implementation of MapReduce is a C++ library with
communication between networked machines via remote procedure calls.
It allows for fault tolerance when large numbers of machines are used,
and can use disks as out-of-core memory to process petabyte-scale data
sets.  Tens of thousands of MapReduce programs have since been written
by Google researchers and are a significant part of the daily compute
tasks run by the company \cite{Dean2}.

Similarly, the open-source Hadoop implementation of MapReduce
\cite{Hadoop}, has become widely popular in the past few years for
parallel analysis of large-scale data sets at Yahoo and other
data-centric companies, as well as in university and laboratory
research groups, due to its free availability.  MapReduce programs in
Hadoop are typically written in Java, though it also supports use of
stand-alone map and reduce kernels, which can be written as shell
scripts or in other languages.

More recently, MapReduce formulations of traditional number-crunching
kinds of scientific computational tasks have been described, such as
post-processing analysis of simulation data \cite{Tu}, graph
algorithmics \cite{Cohen}, and linear algebra operations \cite{Fox}.
The paper by Tu et al \cite{Tu} was particularly insightful to us,
because it described how MapReduce could be implemented on top of the
ubiquitous distributed-memory message-passing interface (MPI), and how
the intermediate data-shuffle operation is conceptually identical to
the familiar $MPI_Alltoall$ operation.  Their implementation of
MapReduce was within a Python wrapper to simplify the writing of user
programs.  The paper motivated us to develop our own C++ library built
on top of MPI for use in graph analytics, which we initially released
as open-source software in mid-2009 \cite{MRMPI}.  We have since
worked to optimize several of the library's underlying algorithms and
to enable its operation in out-of-core mode on larger data sets.
These algorithmic improvements are described in this paper and are
part of the current downloadable version \cite{MRMPI}.

The MapReduce-MPI (MR-MPI) library described in this paper is a
simple, lightweight implementation of basic MapReduce functionality,
with the following features and limitations:

\begin{itemize}

\item {\it C++ library using MPI for inter-processor communication:}
The user writes a (typically) simple main program which runs on each
processor of a parallel machine, making calls to the MR-MPI library.
For map and reduce operations, the library calls back to user-provided
map() and reduce() functions.  The use of C++ allows precise control
over the memory and format of data allocated by each processor during
a MapReduce.  Library calls for performing a map, reduce, or data
shuffle, are synncrhonous, meaning all the processors participate and
finish the operation before proceeding.  Similarly, the use of MPI
within the library is the traditional mode of $MPI_Send()$ and
$MPI_Recv$ calls between processor pairs using large aggregated
messages to improve bandwidth performance and reduce latency costs.  A
recent paper by \cite{Dongarra} also outlines the MapReduce formalism
from an MPI perspective, though they advocate a more asyncrhonous
approach, using one-way communication of small messages.

\item {Small, portable:} The entire MR-MPI library is a few thousand
lines of standard C++ code.  For parallel operation, the program is
linked with MPI, a standard message passing library available on all
distributed memory machines.  For serial operation, a dummy MPI
library (provided) can be substituted.

\item {\it In-core or out-of-core operation:} Each MapReduce object a
processor defines, allocates per-processor "pages" of memory, where
the page size is determined by the user.  Typical MapReduce operations
can be performed using a few such pages.  If the data set fits in a
single page (per processor), then the library performs its operations
in-core.  If the data set exceeds the page size, then processors each
write to temporary disk files (to local disk or a parallel file
system) as needed and subsequently read from them.  This allows
processing of data sets larger than the aggregate memory of all the
processors, i.e. up to the available aggregate disk space.

\item {\it Flexible programmability:} An advantage of writing a
MapReduce program on top of MPI, is that the user program can invoke
MPI calls directly, if desired.  For example, one-line calls to
$MPI_Allreduce$ are often useful in determining the status of an
iterative graph algorithm, as described in Section \ref{sec:graph}.
The library interface also provides a user data pointer as an argument
passed back to all callback functions, so it is easy for the user
program to store ``state'' on each processor, accessible during the
map and reduce operaions.  For example, various flags can be stored
that alter the operation of a map or reduce function, as can richer
data structures, that accumulate the results.

\item {\it C++, C, and Python interfaces:} A C++ interface to the
MR-MPI library means a user program instantiates and then invokes
methods in one or more MapReduce objects.  A C interface means the
library can also be called from C or other hi-level languages such as
Fortran.  A C interface also means the library can be easily wrapped
by Python via the Python ``ctypes'' module.  The library can then be
called from a Python script, allowing the user to write map() and
reduce() callback functions in Python.  If a machine supports running
Python in parallel, a parallel MapReduce can also be run in this mode.

\item {No fault tolerance:} Current MPI implementations do not enable
easy detection of a dead processor, or retrieval of the data it was
working on.  So like most MPI programs, a parallel program calling the
MR-MPI library will hang or crash if a processor goes away.  Unlike
Hadoop, and its HDFS file system which provides for data redundancy,
the MR-MPI library simply reads and writes simple, flat files.  It can
use local per-processor disks, or a parallel file system, if
available, but these typically provide no data redundancy.

\end{itemize}

The remainder of the paper is organized as follows.  The next two
sections \ref{sec:mr} and \ref{sec:outcore} describe how in-core and
out-of-core MapReduce primitives are formulated as MPI-based
operations in the MR-MPI library.  Section \ref{sec:graph} briefly
describes the formulation of several common graph algorithms as
MapReduce operations.  Section \ref{sec:results} gives performance
results for these algorithms running on a parallel cluster for graphs
ranging in size from 1 million to 1 trillion vertices or edges.  In
this section, we highlight the performance and complexity trade-offs
of a MapReduce approach versus other more special-purpose algorithms.
The latter generally perform better but are harder to implement
efficiently on distributed memory machines, due to the required
explicit management of parallelism, particularly for large out-of-core
data sets.  Section \ref{sec:lessons} summarizes some lessons learned
from the implementation and use of our library.

### 2010-03-26 paper/mr.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@479 3fc257bb-8550-0410-81b3-a5c103da83f7

\section{MapReduce in MPI}

@@

>>>>

\section{MapReduce in MPI}
\label{sec:mr}

@@

MR-MPI~\cite{MRMPIManual} is a C++ library using MPI for
inter-processor communication.  The basic data units in MapReduce are
key/value pairs.  The data in keys and values can have any type or
length; they are stored in simple character arrays.  A related data
type is the key/multivalue pair, where all values associated with the
same key are collected and stored together as a multivalue.

The three basic operations in a typical MR-MPI program are {\it
map()}, {\it collate()}, and {\it reduce()}.  The {\it map()}
operation generates key/value pairs on each processor.  Typically,
this is done from data read in from files, but a {\it map()} may
generate data itself or process existing key/value pairs to generate
new ones.  The key/value pairs produced are stored locally on each
processor; a {\it map()} requires no inter-processor communication.
Users provide a pointer to their own map function; the MR-MPI {\it
map()} operation invokes the user function as a callback.

The {\it collate()} operation identifies unique keys and collects all
the values associated with those keys to create key/multivalue pairs.
This is done in two stages, the first of which requires communication,
since key/value pairs with the same key may be owned by any processor.
Each processor hashes its keys to an ``owning'' processor.  It then
sends each of its key/value pairs to the processor which owns the key.
After receiving key/value pairs, the second stage is an on-processor
computation, whereby each processor constructs key/multivalue pairs
for each unique key it now owns, using a local hash table.  Thus the
overall {\it collate()} is equivalent to an all-to-all communication
using a distributed hash table followed by a local reordering of the
data.  We use synchronous point-to-point irregular communication
instead of the MPI\_Alltoall() library call, since this allows control
of message sizes and sequencing.

The {\it reduce()} operation processes key/multivalue pairs and can
produce new key/value pairs for continued computation.  Each processor
operates on the key/multivalue pairs it owns; no communication is
required.  As with the {\it map()}, users provide pointers to their
own reduce functions.  The MR-MPI {\it reduce()} invokes the user's
function once for each key/multivalue pair.  Output from a {\it
reduce()} operation can serve as input to a subsequent {\it map()} or
{\it collate()}.

Several related operations are provided by MR-MPI.  For example, a
{\it compress()} combines on-processor key/value pairs to eliminate
duplicate keys, forming single key/value pairs with an aggregated
value.  The {\it gather()} operation collects key/values pairs from
all processors to a subset of processors; it is useful for doing
output from a single or subset of processors.  Library calls for
sorting data by key or value or within multivalues are also provided.

Since the MapReduce paradigm was designed to enable processing of very
large data sets, we provide some details of how MR-MPI manages memory
and large disk files.  The user (optionally) specifies a {\it memsize}
parameter when instantiating a MapReduce object, e.g. 1 Gbyte.  This
amount of memory is allocated by each processor for storing key/value
and key/multivalue pairs, as well as communication buffers and a hash
table for the {\it collate{}} operation.

If the data generated by various MR-MPI operations can be stored by
each processor within its allocated chunk of memory, we refer to this
as ``in-core'' processing and no disk files are written or read (aside
from initial input data if it exists or final output data if
generated).  Three additional points are worth noting.

At some stages in a MapReduce algorithm, the MapReduce object will
typically need to store both key/value and key/multivalue pairs at the
same time.  For example, during a {\it reduce()} operation,
key/multivalue pairs serve as input to the user's reduce function,
while the user's function may be generating new key/value pairs.

The memory available for storing key/value and key/multivalue pairs is
{\it memsize} summed across all processors.  So long as the key/value
pairs remain reasonably evenly distributed across processors then this
can be a significant amount of memory, e.g. 1 Tbyte on a
1024-processor machine.  The hash function used by the {\it collate()}
operation attempts to randomize key assignment to processors so
effective load-balance occurs.

For some MapReduce algorithms, it is convenient to create multiple
MapReduce objects and pass data back and forth between them.
Obviously, the user must then partition physical memory judiciously
between these various objects if all the data is to stay in-core.  The
{\it memsize} parameter can be specified for each MapReduce object
created.

### 2010-03-26 paper/outcore.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@479 3fc257bb-8550-0410-81b3-a5c103da83f7

\section{Out-of-core Issues}

@@

>>>>

\section{Out-of-core Issues}
\label{sec:outcore}

@@

If the set of key/value or key/multivalue pairs generated or owned by
a processor exceeds the {\it memsize} of its allocated memory, then
MR-MPI switches to ``out-of-core'' processing.  Each processor uses
temporary files to store data that does not fit in its memory.
Depending on the parallel machine's hardware configuration, these
files can reside on disks local to each processor, on the front end
(typically a NSF-mounted file system for a parallel machine), or on a
parallel file system/disk array.  The files are read/written as large
contiguous ``pages'', which are 1/4 the size of the {\it memsize}
setting discussed in the previous section.

We now explain how the MapReduce primitives described above operate in
out-of-core mode.  The {\it map()} and {\it reduce()} operations are
relatively simple.  As a {\it map()} generates key/value pairs, it
fills up a page, one key/value pair at a time.  When the page is full,
it is written to disk.  If the source of data for the {\it map()}
operation is an existing set of key/value pairs, then those datums are
read in, one page at a time, and each key/value pair is given to the
user map() function.  Similarly, a {\it reduce()} reads one page of
key/multivalue pairs from disk, passes them one at a time to the user
reduce() function, which typically generates new key/value pairs.  The
generated pairs fill up a new page, which is written to disk when
full, just as with the {\it map()} operation.  Thus for both a {\it
map()} and {\it reduce()}, out-of-core disk files are read and written
sequentially, one page at a time.

A special case is when a single key/multivalue pair is larger than a
single page.  This can happen, for example, in a connected component
finding algorithm if the graph collapses into one or a few giant
components.  In this case, the entire set of values (graph vertices
and edges) associated with the unique key (the component ID), may not
fit in one page of memory.  The individual values in the multivalue
will then be spread across as many pages as needed.  The user reduce()
function is passed a flag indicating it is receiving only a portion of
the values.  Once it has processed them, it can request a new set of
values (from a new page) from the MR-MPI library.

Performing a {\it collate()} operation out-of-core is more complex.
Recall that a collate() operation occurs in two stages.  First, keys
are hashed to ``owning'' processors and the key/value data is
communicated in an all-to-all fashion.  This is done one page of
key/value data at a time.  Each processor reads in a page of key/value
data, the communication pattern is determined (who needs what datums),
and the communication is performed.  So long as each processor
receives no more than 2x its share of the datums (one page), it has
room to receive it within the {\it memsize} restriction.  If the
key/value data is load-imbalanced so that this limit is exceeded, then
the all-to-all communication can be performed in multiple passes.

When the first stage is done, each processor now has an out-of-core
set of key/value datums that need to be reorganized into an
out-of-core set of key/multivalue pairs.  In principle, to create one
page of key/multivalue data, the entire set of key/value pairs needs
to be scanned to find all keys which contribute values to that page.
Reading all the key/value pages from disk for each page of
key/multivalue output could be prohibitively expensive.  A related
issue is that a hash table is needed to match new keys with previously
encountered keys.  But there is no guarantee that the hash table
itself will fit within the memory specified by the user via the {\it
memsize} parameter.  Our solution to these issues is to partition the
{\it memsize} memory into 3 pieces: 2 pieces each of size {\it
memsize}/4 for reading/writing pages of key/value and key/multivalue
pairs, and the remaining {\it memsize}/2 piece for a hash table.  We
can then scan the key/value pairs at most four times to produce all
the pages of key/multivalue output, as diagrammed in Figure
\ref{f:collate}.

\begin{figure}
\includegraphics[width=\textwidth,angle=-90]{fig_collate.pdf}
\caption{Four-pass algorithm for converting key/value data to
key/multivalue data.  The vertical lines represent out-of-core data
sets.  KV is key/value pairs; HT is an in-memory hash table; KMV is
key/multivalue pairs.}
\label{f:collate}
\end{figure}

\begin{itemize}

\item (Pass 1) The key/values are read and split into two files.  The
first file contains all the key/values pairs whose keys fit into the
finite-size hash table.  The second contains those which do not.

\item (Pass 2) The size of the first file (generally smaller) can be
used to estimate how many chunks the second file (generally larger)
needs to be partitioned into so that each chunk contains a set of
unique keys than can fit into the hash table.  This estimate is made
conservatively so that additional passes through the data are almost
never required.  The second pass thus reads the larger file and writes
out smaller files, similar in size to the first file created by the
first pass.

\item (Pass 3) The unique keys in each small file of key/value pairs
can now be hashed, but the file may contain more value data than will
fit in one key/multivalue page.  As each small file is read, and its
keys hashed, the accumulating hash table statistics can be used to
determine which key/values correspond to single pages of
key/multivalue output.  The key/values can thus be split into even
smaller files, one for each key/multivalue page.

\item (Pass 4) Each smaller file of key/value pairs is read, its keys
hashed, and the key/value data reorganized into sequential
key/multivalue pairs which fill a key/multivalue page, which is then
written out to disk.

\end{itemize}

Note that depending on the data set size and its key distribution all
four passes through the data may not be necessary, but the upper limit
is four, assuming the chunk-size estimations are accurate.  By
comparison, a merge sort of key/value pairs would require $log_2(N)$
passes (read/write) through the out-of-core data set, where $N$ is its
number of pages.  The MR-MPI {\it collate()} operation does not
actually sort the key/multivalue pairs by key; the final ordering is
indeterminate as it is created by the hash function.

The other MR-MPI library calls ({\it compress()}, {\it gather()},
sorting by key or value, etc) can also operate in out-of-core mode,
typically with one pass through the key/value or key/multivalue data.
Sorting is an exception, as an out-of-core merge sort requires
$log_2(N)$ passes, as just mentioned, where $N$ is the number of
pages.

### 2010-03-26 paper/paper.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@479 3fc257bb-8550-0410-81b3-a5c103da83f7

\author{
Steven J.~Plimpton, Karen D.~Devine, Jon?, Cohen?, Others? \\
Sandia National Laboratories \\
Albuquerque, NM \\
sjplimp@sandia.gov
}

@@

We describe a parallel library that allows algorithms to be expressed
in the MapReduce paradigm, where the calling program does not need to
include explicit parallel code, but instead provides ``map'' and
``reduce'' functions which operate in parallel on elements of a data
set distributed across processors.  The library performs needed
parallel operations using message-passing (MPI) calls.  We describe
how typical MapReduce operations can be implemented in an MPI context,
and also in an out-of-core manner for data sets that do not fit within
the aggregate memory of a parallel machine.  Our motivation for
creating this library was to enable graph algorithms to be written as
MapReduce operations, allowing processing of Terabyte-scale data sets.
We outline MapReduce versions of several such algorithms: vertex
ranking via PageRank, triangle finding, connected component
identification, Luby's algorithm for maximally independent sets, and
single-source shortest-path.  To test the algorithms on arbitrarily
large artificial graphs we generate randomized RMAT matrices in
parallel; we also describe a MapReduce version of this operation.
Performance and scalability results for the various algorithms are
presented for graphs of varying sizes on a distributed-memory cluster.
For some cases, we compare the results with non-MapReduce algorithms,
different machines, and different MapReduce software, namely Hadoop.
Our open-source library is written in C++, is callable from C++, C,
Fortran, or scripting languages such as Python, and can run on any
parallel platform that supports MPI.

@@

\input{intro}
\input{mr}
\input{outcore}
\input{graph}
\input{results}
\input{thanks}

>>>>

\author{
Steven J.~Plimpton, Karen D.~Devine, Others? \\
Sandia National Laboratories \\
Albuquerque, NM \\
sjplimp@sandia.gov
}

@@

We describe a parallel library written with message-passing (MPI)
calls that allows algorithms to be expressed in the MapReduce
paradigm.  This means the calling program does not need to include
explicit parallel code, but instead provides ``map'' and ``reduce''
functions which operate independently on elements of a data set
distributed across processors.  The library performs needed data
movement between processors.  We describe how typical MapReduce
funtionality can be implemented in an MPI context, and also in an
out-of-core manner for data sets that do not fit within the aggregate
memory of a parallel machine.  Our motivation for creating this
library was to enable graph algorithms to be written as MapReduce
operations, allowing processing of Terabyte-scale data sets.  We
outline MapReduce versions of several such algorithms: vertex ranking
via PageRank, triangle finding, connected component identification,
Luby's algorithm for maximally independent sets, and single-source
shortest-path calculation.  To test the algorithms on arbitrarily
large artificial graphs we generate randomized RMAT matrices in
parallel; a MapReduce version of this operation is also described.
Performance and scalability results for the various algorithms are
presented for varying size graphs on a distributed-memory cluster.
For some cases, we compare the results with non-MapReduce algorithms,
different machines, and different MapReduce software, namely Hadoop.
Our open-source library is written in C++, is callable from C++, C,
Fortran, or scripting languages such as Python, and can run on any
parallel platform that supports MPI.

@@

\input{intro}
\input{mr}
\input{outcore}
\input{graph}
\input{results}
\input{lessons}
\input{thanks}

### 2010-03-26 paper/thanks.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@479 3fc257bb-8550-0410-81b3-a5c103da83f7

\section{Acknowledgements}

Sandia is a multiprogram laboratory operated by Sandia Corporation, a
Lockheed Martin Company, for the United States Department of
Energy's National Nuclear Security Administration under contract
DE-AC04-94AL85000.

Hadoop folks.

>>>>

\section{Acknowledgements}
\label{sec:thanks}

We thank the following individuals for their contributions to this
paper: Greg Bayer and Todd Plantenga (Sandia) for explaining Hadoop
concepts to us, and for the Hadoop implementations and timings of
Section \ref{sec:results}; Jon Cohen (NSA) for fruitful discussions
about his MapReduce graph algorithms \cite{Cohen}; Brian Barrett
(Sandia) for the PBGL results of Section \ref{sec:results}; Jon Berry
(Sandia) for the MTGL results of Section \ref{sec:results}, and for
his overall support of this work and many useful discussions.

Sandia National Laboratories is a multi-program laboratory operated by
Sandia Corporation, a wholly owned subsidiary of Lockheed Martin
company, for the U.S. Department of Energy's National Nuclear Security
Administration under contract DE-AC04-94AL85000.
